---
title: "Task_2_ByTemplate"
author: "Kasandika Andariefli"
date: "2025-03-06"
output:
  pdf_document: default
  html_document: default
---
---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 2"
output:
  pdf_document: 
    df_print: default
    highlight: tango
    keep_tex: yes
    latex_engine: xelatex
header-includes:
  \usepackage{fvextra}
---

```{r setup, include=FALSE}
library(data.table)
library(ggplot2)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE, linewidth = 80)
```

## Load Data

```{r 0. Load data}
data <- read.csv("QVI_data.csv")
head(data)

# Ensure DATE is in Date format
data$DATE <- as.Date(data$DATE, format="%Y-%m-%d")

# Create YearMonth Column
setDT(data)  # Ensure 'data' is a data.table
data[, YEARMONTH := as.numeric(format(DATE, "%Y%m"))]
```

## Compute Store Metrics

```{r 1. Aggregate Data}
measureOverTime <- data[, .(
  totSales = sum(TOT_SALES),
  nCustomers = uniqueN(LYLTY_CARD_NBR),
  nTxnPerCust = .N / uniqueN(LYLTY_CARD_NBR),
  nChipsPerTxn = sum(PROD_QTY) / .N,
  avgPricePerUnit = sum(TOT_SALES) / sum(PROD_QTY)
), by = .(STORE_NBR, YEARMONTH)]
```

## Control Store Selection

```{r 2. Define Correlation Function}
calculateCorrelation <- function(inputTable, metricCol, storeComparison) {
  calcCorrTable <- data.table(Store1 = numeric(), Store2 = numeric(), corr_measure = numeric())
  storeNumbers <- unique(inputTable[, STORE_NBR])
  
  for (i in storeNumbers) {
    merged_data <- merge(
      inputTable[STORE_NBR == storeComparison, .(YEARMONTH, metric = get(metricCol))],
      inputTable[STORE_NBR == i, .(YEARMONTH, metric = get(metricCol))],
      by = "YEARMONTH", suffixes = c("_trial", "_control"))
    
    if (nrow(merged_data) > 1) {
      corr_value <- cor(merged_data$metric_trial, merged_data$metric_control, use="complete.obs")
    } else {
      corr_value <- NA
    }
    
    calcCorrTable <- rbind(calcCorrTable, data.table(Store1 = storeComparison, Store2 = i, corr_measure = corr_value))
  }
  return(calcCorrTable)
}
```

```{r}
trial_store <- 77
corr_nSales <- calculateCorrelation(measureOverTime, "totSales", trial_store)
corr_nCustomers <- calculateCorrelation(measureOverTime, "nCustomers", trial_store)

# View results
head(corr_nSales)
head(corr_nCustomers)
```


```{r 3. Define Magnitude Distance Function}
calculateMagnitudeDistance <- function(inputTable, metricCol, storeComparison) {
  calcDistTable <- data.table(Store1 = numeric(), Store2 = numeric(), measure = numeric())
  storeNumbers <- unique(inputTable[, STORE_NBR])
  
  for (i in storeNumbers) {
    merged_data <- merge(
      inputTable[STORE_NBR == storeComparison, .(YEARMONTH, metric = get(metricCol))],
      inputTable[STORE_NBR == i, .(YEARMONTH, metric = get(metricCol))],
      by = "YEARMONTH", suffixes = c("_trial", "_control"))
    
    measure_value <- mean(abs(merged_data$metric_trial - merged_data$metric_control), na.rm=TRUE)
    calcDistTable <- rbind(calcDistTable, data.table(Store1 = storeComparison, Store2 = i, measure = measure_value))
  }
  
  minMaxDist <- calcDistTable[, .(minDist = min(measure), maxDist = max(measure)), by = .(Store1)]
  distTable <- merge(calcDistTable, minMaxDist, by = "Store1")
  distTable[, magnitudeMeasure := 1 - (measure - minDist) / (maxDist - minDist)]
  
  return(distTable[, .(Store1, Store2, mag_measure = mean(magnitudeMeasure))])
}
```

```{r}
# Compute magnitude distance for trial store 77
mag_nSales <- calculateMagnitudeDistance(measureOverTime, "totSales", 77)
mag_nCustomers <- calculateMagnitudeDistance(measureOverTime, "nCustomers", 77)

# View results
head(mag_nSales)
head(mag_nCustomers)

```


```{r 4. Compute Control Store Scores}
trial_store <- 77
corr_nSales <- calculateCorrelation(measureOverTime, "totSales", trial_store)
corr_nCustomers <- calculateCorrelation(measureOverTime, "nCustomers", trial_store)
mag_nSales <- calculateMagnitudeDistance(measureOverTime, "totSales", trial_store)
mag_nCustomers <- calculateMagnitudeDistance(measureOverTime, "nCustomers", trial_store)

score_nSales <- merge(corr_nSales, mag_nSales, by = c("Store1", "Store2"))[, scoreNSales := 0.5 * corr_measure + 0.5 * mag_measure]
score_nCustomers <- merge(corr_nCustomers, mag_nCustomers, by = c("Store1", "Store2"))[, scoreNCust := 0.5 * corr_measure + 0.5 * mag_measure]

score_Control <- merge(score_nSales, score_nCustomers, by = c("Store1", "Store2"))
score_Control[, finalControlScore := 0.5 * scoreNSales + 0.5 * scoreNCust]

control_store <- score_Control[Store1 == trial_store][order(-finalControlScore)][2, Store2]
```
```{r}
control_store
```
```{r}
# Prepare data for visualization
pastSales <- measureOverTime[, Store_type := ifelse(
  STORE_NBR == trial_store, "Trial",
  ifelse(STORE_NBR == control_store, "Control", "Other stores"))
][, totSales := mean(totSales), by = c("YEARMONTH", "Store_type")
][, TransactionMonth := as.Date(paste(YEARMONTH %/% 100, YEARMONTH %% 100, 1, sep = "-"), "%Y-%m-%d")
][YEARMONTH < 201903 , ]

# Plot total sales trends
ggplot(pastSales, aes(TransactionMonth, totSales, color = Store_type)) +
  geom_line() +
  labs(x = "Month of operation", y = "Total sales", title = "Total Sales by Month")

```
## Trial Assessment

```{r 5. Compute Trial Uplift}
scalingFactorForControlSales <- measureOverTime[STORE_NBR == trial_store & YEARMONTH < 201902, sum(totSales)] / 
  measureOverTime[STORE_NBR == control_store & YEARMONTH < 201902, sum(totSales)]

scaledControlSales <- measureOverTime[STORE_NBR == control_store, ][, controlSales := totSales * scalingFactorForControlSales]
```

```{r 6. Statistical Testing}
percentageDiff <- merge(scaledControlSales[, .(YEARMONTH, controlSales)], 
                        measureOverTime[STORE_NBR == trial_store, .(YEARMONTH, totSales)], 
                        by = "YEARMONTH")[, percentageDiff := abs(controlSales - totSales) / controlSales]

stdDev <- sd(percentageDiff[YEARMONTH < 201902 , percentageDiff])
percentageDiff[, tValue := percentageDiff / stdDev]
```


```{r 7. Visualization}
library(ggplot2)
library(lubridate)
library(scales)

# Convert YEARMONTH to a proper date format (assuming numeric YYYYMM)
percentageDiff$YEARMONTH <- as.Date(paste0(percentageDiff$YEARMONTH, "01"), format="%Y%m%d")

ggplot(percentageDiff, aes(x = YEARMONTH, y = percentageDiff)) +
  geom_line() +
  labs(title = "Trial Uplift Analysis", x = "Month", y = "Percentage Difference") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Fluctuations in the percentage difference over time. There are peaks in September 2018 and February 2019, dips are in August 2018 and November 2018 as well as a sharp drop in March 2019 after the February peak followed by a gradual increase in April 2019.

* Volatility appears too high, external factors may have heavily influenced the trial’s success.
* Sustaining uplift beyond peak months is a challenge.
* Investigate what happened in September 2018, February 2019, and March 2019 to understand the causes behind the fluctuations.
