---
title: "Quantum_Task1"
author: "Kasandika Andariefli"
date: "2025-03-04"
output:
  pdf_document: default
  html_document: default
---


```{r}
library(readr)      
library(readxl)     
library(dplyr)     
library(ggplot2)    
```


# Data Loading
```{r}
purchase_data <- read_csv("QVI_purchase_behaviour.csv")
transaction_data <- read_excel("QVI_transaction_data.xlsx", sheet = 1)

```
```{r}
str(purchase_data )
```
```{r}
head(purchase_data$LYLTY_CARD_NBR)
head(purchase_data$LIFESTAGE)
head(purchase_data$PREMIUM_CUSTOMER)
```

```{r}
str(transaction_data)
```
DATE should be in a proper Date format instead of numeric
```{r}
transaction_data$DATE <- as.Date(transaction_data$DATE, origin = "1899-12-30")
str(transaction_data)  # DATE should now be Date type
head(transaction_data$DATE) 
```
```{r}
head(purchase_data)
head(transaction_data)
```
```{r}
summary(purchase_data)
summary(transaction_data)
```
## Checking for missing value
```{r}
colSums(is.na(purchase_data))
colSums(is.na(transaction_data))
```
## Checking for duplicates
```{r}
sum(duplicated(purchase_data))
sum(duplicated(transaction_data))
```
```{r}
# Removing duplicated values
transaction_data <- transaction_data %>% distinct()
sum(duplicated(transaction_data))
```

# Feature Engineering
```{r}
head(transaction_data$PROD_NAME)
```
The pack size is at the end of the PROD_Name, in the ...g format, we aim to extract the numeric value before the "g".
```{r}
# transaction_data$PACK_SIZE <- as.numeric(gsub(".*?(\\d{2,3})g$", "\\1", transaction_data$PROD_NAME))
#vtransaction_data$PACK_SIZE <- as.numeric(gsub(".*?(\\d{2,3})[gG]$", "\\1", trimws(transaction_data$PROD_NAME)))
transaction_data$PACK_SIZE <- as.numeric(gsub(".*?(\\d{2,3})[gG].*", "\\1", trimws(transaction_data$PROD_NAME)))
```
```{r}
sum(is.na(transaction_data$PACK_SIZE))
transaction_data[is.na(transaction_data$PACK_SIZE), "PROD_NAME"]
```

Extracting the name of the pack
```{r}
transaction_data$BRAND <- gsub("^(\\S+).*", "\\1", transaction_data$PROD_NAME)
```
```{r}
head(transaction_data[, c("PROD_NAME", "BRAND", "PACK_SIZE")])
```

# Exploratory Data Analysis

## Total Spend per Customer Segment
```{r}
library(ggplot2)
library(dplyr)

library(dplyr)

customer_spend <- transaction_data %>%
  left_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(TOTAL_SPEND = sum(TOT_SALES, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_SPEND))

print(customer_spend)

# Compute total spend
customer_spend <- transaction_data %>%
  left_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(TOTAL_SPEND = sum(TOT_SALES, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_SPEND))

# Plot total spend
ggplot(customer_spend, aes(x = reorder(LIFESTAGE, -TOTAL_SPEND), y = TOTAL_SPEND, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Spend per Customer Segment",
       x = "Lifestage",
       y = "Total Spend ($)",
       fill = "Premium Customer") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Insights

1. Older Singles/Couples and Retirees are among the highest spenders.  
   This is very important.
2. New Families have the lowest spend across all premium tiers.
3. Premium customers are less prominent throughout all customer segments.

Potential Strategy

1. Target high-spending segments (Older Singles/Couples, Retirees, Older Families).
2. Encourage higher spending in younger demographics (Young Singles/Couples, Young Families, New Families).
3. Explore premium product promotions to increase engagement.

## Average Spend per Transaction Analysis
```{r}
avg_spend_txn <- transaction_data %>%
  left_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(AVG_SPEND_TXN = mean(TOT_SALES, na.rm = TRUE)) %>%
  arrange(desc(AVG_SPEND_TXN))

print(avg_spend_txn)

library(ggplot2)


ggplot(avg_spend_txn, aes(x = reorder(LIFESTAGE, -AVG_SPEND_TXN), y = AVG_SPEND_TXN, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Spend per Transaction by Customer Segment",
       x = "Lifestage",
       y = "Average Spend per Transaction ($)",
       fill = "Premium Customer") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Insights
Average spend per transaction is fairly consistent across all customer segments.

## Pack Size Influence on Spending
```{r}
pack_size_spend <- transaction_data %>%
  left_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(PACK_SIZE) %>%
  summarise(AVG_SPEND = mean(TOT_SALES, na.rm = TRUE),
            TOTAL_SPEND = sum(TOT_SALES, na.rm = TRUE),
            TRANSACTIONS = n()) %>%
  arrange(desc(AVG_SPEND))

print(head(pack_size_spend, 10))  # Show top 10 pack sizes by avg spend

# ====
library(ggplot2)

ggplot(pack_size_spend, aes(x = PACK_SIZE, y = AVG_SPEND)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Pack Size vs. Average Spend",
       x = "Pack Size (g)",
       y = "Average Spend ($)") +
  theme_minimal()
```
Insight

1. Positive correlation between pack size and average spend
2. Smaller pack sizes have a wider variance in spend

Potential Strategy

1. Promote Bulk Buying, offer discounts on larger pack sizes to encourage customers to purchase more at once.
2. Optimize Shelf Space & Inventory for Larger Packs

## Purchase Quantity Influence on Spend
```{r}
qty_spend <- transaction_data %>%
  left_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(PROD_QTY) %>%
  summarise(AVG_SPEND = mean(TOT_SALES, na.rm = TRUE),
            TOTAL_SPEND = sum(TOT_SALES, na.rm = TRUE),
            TRANSACTIONS = n()) %>%
  arrange(desc(AVG_SPEND))

print(head(qty_spend, 10))  # Show top 10 purchase quantities by avg spend

library(ggplot2)

ggplot(qty_spend, aes(x = PROD_QTY, y = AVG_SPEND)) +
  geom_point(color = "green", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Purchase Quantity vs. Average Spend",
       x = "Purchase Quantity",
       y = "Average Spend ($)") +
  theme_minimal()
```

Insight

1. Higher Purchase Quantity Leads to Lower Average Spend per Unit, single-unit purchase has an average spend of $3.71, while a 200-unit purchase averages $650 per unit, likely an outlier.
2. Higher Quantities (3-5 Units) Have Strong Spend but Lower Frequency

Potential Strategy: Loyalty Programs for Bulk Buyers: buying 4-5 packs per transaction could benefit from loyalty incentives.


## Brand Popularity Analysis
```{r}
library(ggplot2)
library(dplyr)

brand_popularity <- transaction_data %>%
  group_by(BRAND) %>%
  summarise(TRANSACTIONS = n()) %>%
  arrange(desc(TRANSACTIONS))

ggplot(brand_popularity, aes(x = reorder(BRAND, TRANSACTIONS), y = TRANSACTIONS, fill = BRAND)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top Brands by Transaction Count (Ascending Order)",
       x = "Brand",
       y = "Number of Transactions") +
  theme_minimal() +
  theme(legend.position = "none")

head(brand_popularity, 10)  # Show the top 10 most purchased brands
```
Insights

1. Kettle being significantly ahead of others
2. Smiths, Pringles, and Doritos indicate a strong consumer preference for mainstream, well-known chip brands.
3. Niche brands like French, NCC, and Sunbites have the lowest transaction counts, potentially indicating a smaller or specialized audience.

Potential Strategy

1. High performers such as the top 4 can continue leveraging strong brand loyalty with promotions and limited-time offers.
2. Mid tier brands can benefit from increased marketing and bundle offers to push them into the top-tier segment.
3. Low performers like French, NCC, and Sunbites could use better shelf placement, discounts, or co-marketing with retailers to improve transactions.

## Total Revenue Contribution per Brand
```{r}
brand_revenue <- transaction_data %>%
  group_by(BRAND) %>%
  summarise(TOTAL_REVENUE = sum(TOT_SALES, na.rm = TRUE)) %>%
  arrange(desc(TOTAL_REVENUE))

ggplot(brand_revenue, aes(x = reorder(BRAND, TOTAL_REVENUE), y = TOTAL_REVENUE, fill = BRAND)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top Brands by Total Revenue",
       x = "Brand",
       y = "Total Revenue ($)") +
  theme_minimal() +
  theme(legend.position = "none")

head(brand_revenue, 10)  # Show the top 10 brands with the highest total revenue

```
Insights

1. Kettle leads in total revenue, reinforcing its strong market dominance.
2. The top 4 generates high revenue overall, aligning with their high transaction counts.
3. Few mid tier brands (Old, Thins, Tostitos, and Twisties) perform well in revenue despite lower transaction counts, indicating higher price points or bulk purchases.
4. Low-revenue brands (Sunbites, Burger, French, NCC) may struggle due to lower transaction counts or smaller pack sizes.

Potential Strategy

1. Leverage Pricing Strategy for Mid-Tier Brands by promoting larger pack sizes or premium variants.
2. Maintain a competitive edge for high-performers by branding and targeted promotions.
3. Boost low-performing brands by better visibility, competitive pricing, or bundling with popular brands to increase market share.

## Comparing Brands by Revenue-to-Transaction Ratio
```{r}
brand_analysis <- transaction_data %>%
  group_by(BRAND) %>%
  summarise(
    TRANSACTIONS = n(), 
    TOTAL_REVENUE = sum(TOT_SALES, na.rm = TRUE)
  ) %>%
  mutate(REVENUE_PER_TXN = TOTAL_REVENUE / TRANSACTIONS) %>%
  arrange(desc(REVENUE_PER_TXN))

head(brand_analysis, 10)  # Show the top 10 brands

ggplot(brand_analysis, aes(x = reorder(BRAND, REVENUE_PER_TXN), y = REVENUE_PER_TXN, fill = BRAND)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Brands Ranked by Revenue per Transaction",
       x = "Brand",
       y = "Revenue per Transaction ($)") +
  theme_minimal() +
  theme(legend.position = "none")

```
Potential Strategy
1. Brands with high revenue per transaction (e.g., Dorito, Old, Kettle) can explore bundling options to encourage repeat purchases.
2. Low-performing brands should consider larger pack sizes or premium variants to increase revenue per transaction.

## Average Spend per Transaction per Customer Segment
```{r}
avg_spend_segment <- transaction_data %>%
  left_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(
    TRANSACTIONS = n(),
    TOTAL_SPEND = sum(TOT_SALES, na.rm = TRUE),
    AVG_SPEND_PER_TXN = TOTAL_SPEND / TRANSACTIONS
  ) %>%
  arrange(desc(AVG_SPEND_PER_TXN))

print(avg_spend_segment)

ggplot(avg_spend_segment, aes(x = reorder(interaction(LIFESTAGE, PREMIUM_CUSTOMER), AVG_SPEND_PER_TXN), 
                              y = AVG_SPEND_PER_TXN, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Average Spend per Transaction by Customer Segment",
       x = "Customer Segment",
       y = "Average Spend per Transaction ($)") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

## Total Spend Breakdown: LIFESTAGE vs. PREMIUM_CUSTOMER
```{r}
library(scales)  # For formatting numbers

ggplot(avg_spend_segment, aes(x = LIFESTAGE, y = TOTAL_SPEND, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Total Spend by Customer Segment",
       x = "Lifestage",
       y = "Total Spend ($)",
       fill = "Premium Customer Tier") +
  scale_y_continuous(labels = scales::comma) +  # Ensures full number formatting
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Insights

1. Older Singles/Couples and Retirees spend the most overall.
2. Budget customers dominate the top-spending groups.
3. New Families have the lowest spend, with fewer Premium/Mainstream customers.
4. Underage and Young Singles/Couples Have Moderate Spending

Potential Strategy

1. Targeting Older Singles/Couples & Retirees by promoting premium tier products or offering loyalty programs
2. Engage with New Families and Underage Singles/Couples by introducing affordable bulk-pack options to attract budget-conscious buyers
3. Enhancing brand appeal for Young Singles & Couples 

## Brand Preferences by Customer Segment
```{r}
library(ggplot2)
library(dplyr)

# Summarizing total spend per brand and customer segment
brand_spend <- transaction_data %>%
  inner_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(BRAND, LIFESTAGE) %>%
  summarise(TOTAL_SPEND = sum(TOT_SALES), .groups = 'drop')

# Plot
ggplot(brand_spend, aes(x = reorder(BRAND, -TOTAL_SPEND), y = TOTAL_SPEND, fill = LIFESTAGE)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Spend per Brand by Customer Segment",
       x = "Brand",
       y = "Total Spend ($)",
       fill = "Lifestage") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10))

# Finding top 5 brands per customer segment
top_brands <- transaction_data %>%
  inner_join(purchase_data, by = "LYLTY_CARD_NBR") %>%
  group_by(LIFESTAGE, BRAND) %>%
  summarise(TOTAL_SPEND = sum(TOT_SALES), .groups = 'drop') %>%
  arrange(desc(TOTAL_SPEND)) %>%
  group_by(LIFESTAGE) %>%
  slice_max(order_by = TOTAL_SPEND, n = 5)

# Plot
ggplot(top_brands, aes(x = reorder(BRAND, -TOTAL_SPEND), y = TOTAL_SPEND, fill = LIFESTAGE)) +
  geom_bar(stat = "identity") +
  facet_wrap(~LIFESTAGE, scales = "free_x") +
  labs(title = "Top 5 Brands Preferred by Each Customer Segment",
       x = "Brand",
       y = "Total Spend ($)",
       fill = "Lifestage") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
Insights

1. Kettle and Smiths are the top brands in terms of spending.
2. Most segments have a diverse spending pattern, but some brands dominate significantly.
3. Kettle is the most preferred brand across all customer segments.
4. Smiths, Doritos, and Pringles continue to maintain their high performance.

Potential Strategy

1. Retirees and Older Families prefer Kettle, Smiths, and Doritos, consider bundling these brands together for promotions.
2. Young Singles/Couples and Young Families also spend on similar brands but at lower amounts, engage them with budget-friendly promotions.
3. Older Singles/Couples and Older Families, as they contribute the most to total spending, offer loyalty programs, bundle deals, or personalized discounts for these groups.
4. Strengthen partnerships with the top 4 brands (Kettle, Smiths, Doritos, and Pringles) as they dominate across all segments
5. Ensure Kettle, Smiths, Doritos, and Pringles have optimal stock levels across all stores.

## Popular Pack Size Analysis
```{r}
library(ggplot2)
library(dplyr)

# Total Spend per Pack Size
total_spend_packsize <- transaction_data %>%
  group_by(PACK_SIZE) %>%
  summarise(TOTAL_SPEND = sum(TOT_SALES)) %>%
  arrange(desc(TOTAL_SPEND))

ggplot(total_spend_packsize, aes(x = reorder(PACK_SIZE, -TOTAL_SPEND), y = TOTAL_SPEND)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Spend by Pack Size", x = "Pack Size (g)", y = "Total Spend ($)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Most Frequently Purchased Pack Sizes
freq_packsize <- transaction_data %>%
  group_by(PACK_SIZE) %>%
  summarise(TRANSACTIONS = n()) %>%
  arrange(desc(TRANSACTIONS))

ggplot(freq_packsize, aes(x = reorder(PACK_SIZE, -TRANSACTIONS), y = TRANSACTIONS)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Most Frequently Purchased Pack Sizes", x = "Pack Size (g)", y = "Transaction Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Insights: 175g, 150g, and 134g packs dominate in both total spend and transaction count.

Potential Strategy

1. Prioritize stocking 175g, 150g, and 134g packs as they drive the most sales.
2. Offer bulk discounts on 300g+ packs to encourage larger purchases.
3. Investigate if customers prefer 175g and 150g due to pricing vs. value perception.

## Average Spend per Pack Size.
```{r}
# Average Spend per Pack Size
avg_spend_packsize <- transaction_data %>%
  group_by(PACK_SIZE) %>%
  summarise(AVG_SPEND = mean(TOT_SALES)) %>%
  arrange(desc(AVG_SPEND))

ggplot(avg_spend_packsize, aes(x = reorder(PACK_SIZE, -AVG_SPEND), y = AVG_SPEND)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Average Spend per Pack Size", x = "Pack Size (g)", y = "Average Spend ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Insights

1. Larger pack sizes (380g, 330g, 270g) have the highest average spend per transaction.
2. Mid-range packs (150g, 175g, 134g, 110g) have moderate but consistent average spend.
3. Smaller pack sizes (90g, 160g, 200g) have the lowest average spend.

Potential Strategy

1. Since large packs drive higher spending, consider premium pricing or bundling offers.
2. Maintain Focus on Popular Sizes (175g, 150g, 134g, 110g)
3. If profitability is low, consider discontinuing smaller packs (<125g) or repositioning them as trial/sample products.


## Transaction Count per Pack Size
```{r}
# Transaction Count per Pack Size
txn_count_packsize <- transaction_data %>%
  group_by(PACK_SIZE) %>%
  summarise(TRANSACTION_COUNT = n()) %>%
  arrange(desc(TRANSACTION_COUNT))

ggplot(txn_count_packsize, aes(x = reorder(PACK_SIZE, -TRANSACTION_COUNT), y = TRANSACTION_COUNT)) +
  geom_bar(stat = "identity", fill = "orange") +
  labs(title = "Transaction Count per Pack Size", x = "Pack Size (g)", y = "Transaction Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Insights

1. 175g, 150g, and 134g packs dominate transactions.
2. Smaller packs (e.g., 125g, 180g, 160g, 220g) have minimal transactions.
3. Larger packs (e.g., 380g, 330g, 300g) have lower transaction counts.

Potential Strategies

1. Maintain strong stock of 175g, 150g, and 134g packs.
2. Bundle the mid range pack sizes with discounts or loyalty rewards to increase sales.
3. Reassess the viability of smaller packs (<125g) due to low value.


```{r}
# Ensure DATE column is in Date format
transaction_data$DATE <- as.Date(transaction_data$DATE, origin = "1899-12-30")

# Aggregate total sales per month
monthly_sales <- transaction_data %>%
  mutate(MONTH = as.Date(cut(DATE, breaks = "month"))) %>%
  group_by(MONTH) %>%
  summarise(TOTAL_SALES = sum(TOT_SALES))

ggplot(monthly_sales, aes(x = MONTH, y = TOTAL_SALES)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Monthly Sales Trend", x = "Month", y = "Total Sales ($)") +
  theme_minimal()

```
Insights

1. Highest in Jan 2019 and Apr 2019.
2. Sharp drop in Feb 2019, suggesting possible seasonal or external factors.
3. Sales remain relatively stable between 155,000 - 170,000
4. Sales drop significantly in Feb 2019, potentially due to post-holiday demand drop.

Potential Strategies

1. Investigate Feb 2019 drop: Was there a supply chain issue, reduced demand, or pricing changes?
2. Capitalize on peak months: Reinforce promotions before January and April.